---
title: "ST2195 Coursework"
author: "ylei003"
date: "2023-02-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set working directory to the source file location
The source file location includes this .Rmd file and dataverse_files
```{r}

```

# Install and load packages
```{r}
# install.packages("dplyr")
library(dplyr)   #Used to manipulate the data frames
#install.packages("tidyr")
library(tidyr)
#install.packages("ggtext")
library(ggtext)
library(ggplot2)
```

## Data Preparation
1. Extract all compressed flight data files in the "dataverse_files" folder
2. Read the extracted .csv files into data frames
```{r}
df_2005 <- read.csv("dataverse_files/2005.csv", header = TRUE, sep = ',') 
df_2006 <- read.csv("dataverse_files/2006.csv", header = TRUE, sep = ',') 
# View the first 6 rows
head(df_2005)
head(df_2006)
```

3. Merge the flight data frames
```{r}
flight_df <- df_2005 %>% full_join(df_2006) 
```

4. View the summary and structure of the data frame
```{r}
summary(flight_df)
str(flight_df)
```

5. Remove rows with at least one null value, and check for any duplicated rows
```{r}
flight_df <- flight_df %>% drop_na()
sum(is.na(flight_df))
flight_df %>% distinct()
```

## Data Analysis
# 1. When is the best time of day, day of the week, and time of year to fly to  minimise delays?
```{r}
# Group time into categories
flight_delayed <- filter(flight_df, ArrDelay > 0)
```

Create "DepHour" column in flight_delayed data frame.
```{r}
flight_delayed$DepHour <- trunc(flight_delayed$CRSDepTime/100)

# DepHour contains integer values representing hour of the day (0 ~ 23)
```

# 1a. Best time of day to fly to minimise delays
```{r}
flight_delayed_hour <- flight_delayed %>% group_by(DepHour) %>% summarise(avg_delay = mean(ArrDelay))

flight_delayed_hour
```

```{r}
ggplot(flight_delayed_hour, aes(x = DepHour, y = avg_delay)) +
  geom_line() + 
  ylim(10, 45) +
  scale_x_continuous(breaks = seq(0, 23, 1)) +
  labs(title = "Relationship between Departure Time and Delay", x = "Departure Hour", y = "Average Delay in Minutes")

```

# 1b. Best day of week to fly to minimise delays
```{r}
flight_delayed_day <- flight_delayed %>% group_by(DayOfWeek) %>% summarise(avg_delay = mean(ArrDelay))

flight_delayed_day

```

```{r}
ggplot(flight_delayed_day, aes(x = DayOfWeek, y = avg_delay)) +
  geom_line() + 
  ylim(20, 40) +
  scale_x_continuous(breaks = seq(0, 7, 1)) +
  labs(title = "Relationship between Departure Day and Delay", x = "Day of Week", y = "Average Delay in Minutes")

```

# 1c. Best month of year to fly to minimise delays
```{r}
flight_delayed_month <- flight_delayed %>% group_by(Month) %>% summarise(avg_delay = mean(ArrDelay))

flight_delayed_month

```

```{r}
ggplot(flight_delayed_month, aes(x = Month, y = avg_delay)) +
  geom_line() + 
  ylim(20, 40) +
  scale_x_continuous(breaks = seq(0, 12, 1)) +
  labs(title = "Relationship between Departure Month and Delay", x = "Month", y = "Average Delay in Minutes")

```


# 2. Do older planes suffer more delays?
```{r}
planes <- read.csv("dataverse_files/plane-data.csv", sep=",", header=TRUE)

main_df <- dplyr::left_join(x=flight_df, y=planes[ , c("tailnum", "year")], by = c('TailNum'='tailnum'))

# Join "flight_df" with a subset of the "planes" data frame. The subset contains the "tailnum" and "year" columns. The join is performed by matching the "TailNum" column in "flight_df" with the "tailnum" column in the "planes" data frame. The resulting data frame is called "main_df".
```

```{r}
# Calculate age of plane by using the year of the flight(2005, 2006) minus the manufacture year of the plane. 
main_df$PlaneAge <- main_df$Year - as.integer(main_df$year)

# drop null values and filter out data where year = 0
main_df <- main_df %>% drop_na() %>% filter(year!=0)

str(main_df)
```

```{r}
plane_delay <- main_df %>% group_by(PlaneAge) %>% filter(PlaneAge > 0, PlaneAge < 2005) %>% summarise(avg_delay = mean(ArrDelay))

plane_delay
```

```{r, fig.width=14, fig.height=7}
ggplot(plane_delay, aes(x = PlaneAge, y = avg_delay)) +
  geom_line() + 
  scale_x_continuous(breaks = seq(0, 50, 1)) +
  labs(title = "Relationship between plane age and delay", x = "Plane Age", y = "Average Delay in Minutes")
```


# 3. How does the number of people flying between different locations change over time? 
```{r}
airports <- read.csv("dataverse_files/airports.csv", sep=",", header=TRUE)

main_df <- dplyr::left_join(x=main_df, y=airports[ , c("iata", "airport", "lat", "long")], by = c('Origin'='iata'))

```

Calculate the flight frequencies between different flight routes.
```{r}
flight_freq <- main_df %>%
  left_join(airports, by = c("Origin" = "iata")) %>%  
  select(Year, Origin, Dest) %>%
  group_by(Year, Origin, Dest) %>% 
  count(Origin) %>% 
  arrange(desc(n))

```

Filter top 10 popular routes in 2005.
```{r}
top10_2005 <- flight_freq[flight_freq$Year == 2005, ] %>% 
  arrange(desc(n)) %>%
  mutate(Route = paste0(Origin,' to ', Dest)) %>%
  head(10) %>%
  arrange(desc(Route))

top10_2005

```

Filter to match the top 10 routes in 2005 from 2006 flight data.
```{r}
top10_2006 <- flight_freq[flight_freq$Year == 2006,] %>%
  mutate(Route = paste0(Origin,' to ', Dest)) %>%
  filter(Route %in% top10_2005$Route) %>%
  arrange(desc(Route))

top10_2006

# %in% operator is used to test whether a vector or set of values is contained in another vector or set of values. It returns a logical vector of the same length as the left-hand side vector indicating whether each element of the left-hand side vector is found in the right-hand side vector.
```

Calculate the change in number of flights for 2005 top 10 routes.
```{r}
Route_diff <- top10_2006$n - top10_2005$n

Route_diff
```

```{r}
ggplot(top10_2006, aes(x = top10_2006$Route, y = Route_diff, fill =  Route_diff > 0)) + 
  geom_col() + 
  scale_fill_manual(values = c("lightblue", "beige")) +
  coord_flip() +
  labs(x = "Top 10 flight routes in 2005", y = "Value") +
  ggtitle("Flight change between 2005 and 2006") +
  geom_text(aes(label = signif(Route_diff)), nudge_y = 40, size = 3) +
  theme_minimal()

```


# 4. Can you detect cascading failures as delays in one airport create delays in others?
```{r}
Late_aircraft_delay <- flight_df %>%
                       filter(TailNum != 0 & LateAircraftDelay < 120 & LateAircraftDelay > 0) %>%
                       select(Year, Month, DayofMonth, DepDelay, ArrDelay, 
                              TailNum, Origin, Dest, LateAircraftDelay) %>%
                       arrange(desc(LateAircraftDelay)) %>%
                       rename("Year" = 1, "Month" = 2, "DayofMonth" = 3, "DepDelay" = 4, "ArrDelay" = 5, 
                              "TailNum" = 6, "Origin" = 7, 
                              "Dest" = 8, "LateAircraftDelay" = 9)

```


```{r}
N462UA <- flight_df %>%
          filter(Year == 2005 & Month == 1 & DayofMonth == 3 & TailNum == 'N462UA') %>%
          select(Year, Month, DayofMonth, CRSDepTime, DepDelay, ArrDelay, 
                              TailNum, Origin, Dest, LateAircraftDelay) %>%
          arrange(CRSDepTime) %>%
          rename("Year" = 1, "Month" = 2, "DayofMonth" = 3, "CRSDepTime" = 4, "DepDelay" = 5, "ArrDelay" = 6, 
                 "TailNum" = 7, "Origin" = 8, 
                 "Dest" = 9, "LateAircraftDelay" = 10)

N462UA
```


# 5. Use the available variables to construct a model that predicts delays.
```{r}
library('mlr3')
library('mlr3learners')
library('mlr3pipelines')

```

From the Python results, we found out 6 variables correlated with ArrDelay.
```{r}
modelling_df <- main_df %>%
  select(DepDelay, LateAircraftDelay, CarrierDelay, NASDelay, TaxiOut, WeatherDelay, ArrDelay) %>%
  rename("DepDelay" = DepDelay,
         "LateAircraftDelay" = LateAircraftDelay,
         "CarrierDelay" = CarrierDelay,
         "NASDelay" = NASDelay,
         "TaxiOut" = TaxiOut,
         "WeatherDelay" = WeatherDelay,
         "ArrDelay" = ArrDelay)

```


```{r}
task <- TaskRegr$new('Arr_Delay', backend = modelling_df, target = 'ArrDelay')

measure <- msr("regr.mse")
```

Linear regression model
```{r}
learner_lm <- lrn('regr.lm')
gr_lm <- po('imputemean') %>>%
 po(learner_lm)
glrn_lm <- GraphLearner$new(gr_lm)

```

```{r}
set.seed(123)
train_set <- sample(task$nrow, 0.7 * task$nrow) 
test_set <- setdiff(seq_len(task$nrow), train_set)
glrn_lm$train(task, row_ids = train_set)
glrn_lm$predict(task, row_ids = test_set)$score() 

measure

```


Ridge regression model
```{r}
# install.packages("glmnet")
library(glmnet)

learner_ridge <- lrn('regr.glmnet') 
learner_ridge$param_set$values <- list(alpha = 0)
gr_ridge <- po('scale') %>>%
 po('imputemean') %>>%
 po(learner_ridge)
glrn_ridge <- GraphLearner$new(gr_ridge)

```

```{r}
library('mlr3tuning')
library('paradox')

# Set up tuning environment
tune_lambda <- ParamSet$new (list(
ParamDbl$new('regr.glmnet.lambda', lower = 0.001, upper = 1)
))
tuner<-tnr('grid_search')
terminator <- trm('evals', n_evals = 20)

#Put everything together in a new learner
at_ridge <- AutoTuner$new(
 learner = glrn_ridge,
 resampling = rsmp('cv', folds = 3),
 measure = measure,
 search_space = tune_lambda,
 terminator = terminator,
 tuner = tuner
)

#Train the learner on the training data
at_ridge$train(task, row_ids = train_set)

```


```{r}
at_ridge$predict(task, row_ids = test_set)$score() 

measure

```


Lasso regression model
```{r}
learner_lasso <- lrn('regr.glmnet') 
learner_lasso$param_set$values <- list(alpha = 1, lambda = 0.03)
gr_lasso <- po('scale') %>>%
 po('imputemean') %>>%
 po(learner_lasso)
glrn_lasso <- GraphLearner$new(gr_lasso)
glrn_lasso$train(task, row_ids = train_set)
glrn_lasso$predict(task, row_ids = test_set)$score() 

measure

```








